### 2019-06-05 1-Preamble/5-Abstract/resume_fr.tex
Changed \paragraph{XXX} to \paragraph{XXX.}
Fixed typos in some references

\TODOL{Finir de traduire}
\begin{itemize}
    \item
    Nous présentons les concepts et les notations du problème des bandits à armes multiples dans le chapitre~\ref{chapter:2}, d'un point de vue mathématique.
    Mais nous suivons également une approche didactique puisque nous utilisons une démonstration interactive en ligne conçue pour permettre à n'importe qui de jouer contre un petit problème de bandit depuis son navigateur, dans Section~\ref{par:2:interactiveDemoDemoDiscoverMAB}.

@@

\begin{figure}[h!]
    \centering
    \resizebox{0.95\textwidth}{!}{
    \begin{tikzpicture}[>=latex',line join=bevel,scale=2.25]
        %
        \node[align=center] (introduction) at (0,3.25) [rectangle,draw] {Chapitre~\ref{chapter:1}\\Introduction};
        \node[align=center] (chapter2) at (0,2.25) [rectangle,draw] {Chapitre~\ref{chapter:2}\\Modèle bandit à plusieurs bras\\stochastique et stationnaire};
        \node[align=center] (chapter3) at (2.5,2.25) [rectangle,draw] {Chapitre~\ref{chapter:3}\\SMPyBandits : bibliothèque de\\simulation pour MAB};
        \node[align=center] (chapter4) at (-2.5,1) [rectangle,draw] {Chapitre~\ref{chapter:4}\\Deux modèles MAB\\pour les réseaux IoT};
        \node[align=center] (chapter5) at (0,1) [rectangle,draw] {Chapitre~\ref{chapter:5}\\Modèle MAB\\Multi-joueurs};
        \node[align=center] (chapter6) at (2.5,1) [rectangle,draw] {Chapitre~\ref{chapter:6}\\Modèle MAB\\Non-stationnaire};
        \node[align=center] (conclusion) at (0,-0.25) [rectangle,draw] {Chapitre~\ref{chapter:conclusion}\\Conclusion générale};
        %
        \draw [color=black,thick,->] (introduction) to (chapter2);
        \draw [color=black,thick,<->] (chapter2) to (chapter3);
        \draw [color=black,thick,->] (chapter2) to (chapter4);
        \draw [color=black,thick,->] (chapter2) to (chapter5);
        \draw [color=black,densely dotted,<->]   (chapter4) to (chapter5);
        \draw [color=black,densely dotted,->] -| (chapter3) to (chapter6);
        \draw [color=black,densely dotted,<->]   (chapter5) to (chapter6);
        \draw [color=black,thick,->] (chapter2) to (chapter6);
        \draw [color=black,thick,->] (chapter4) to (conclusion);
        \draw [color=black,thick,->] (chapter5) to (conclusion);
        \draw [color=black,thick,->] (chapter6) to (conclusion);
        %
    \end{tikzpicture}
    }
    \caption{Organisation de la thèse : une carte de lecture. Tout chemin contenant le Chapitre~\ref{chapter:1}, le Chapitre~\ref{chapter:2}, au moins un des trois Chapitres~\ref{chapter:4}, \ref{chapter:5} et \ref{chapter:6}, et la Conclusion est une façon valide de lire cette thèse.}
    \label{fig:1:organization}
\end{figure}

\TODOL{Finir de traduire}
L'ordre de lecture du manuscrit peut être n'importe quel chemin descendant entre l'introduction dans le chapitre actuel~\ref{chapter:1}, et la conclusion dans le dernier chapitre~\ref{chapter:conclusion}.
Comme le montre le graphique de la Figure~\ref{fig:1:organisation},
la thèse est organisée en deux parties :

>>>>

\TODOL{Finir de traduire}
\begin{itemize}
    \item
    Nous présentons les concepts et les notations du problème des bandits à armes multiples dans le chapitre~\ref{chapter:2}, d'un point de vue mathématique.
    Mais nous suivons également une approche didactique puisque nous utilisons une démonstration interactive en ligne conçue pour permettre à n'importe qui de jouer contre un petit problème de bandit depuis son navigateur, dans Section~\ref{par:2:interactiveDemoDiscoverMAB}.

@@

\begin{figure}[h!]
    \centering
    \resizebox{0.95\textwidth}{!}{
    \begin{tikzpicture}[>=latex',line join=bevel,scale=2.25]
        %
        \node[align=center] (introduction) at (0,3.25) [rectangle,draw] {Chapitre~\ref{chapter:1}\\Introduction};
        \node[align=center] (chapter2) at (0,2.25) [rectangle,draw] {Chapitre~\ref{chapter:2}\\Modèle bandit à plusieurs bras\\stochastique et stationnaire};
        \node[align=center] (chapter3) at (2.5,2.25) [rectangle,draw] {Chapitre~\ref{chapter:3}\\SMPyBandits : bibliothèque de\\simulation pour MAB};
        \node[align=center] (chapter4) at (-2.5,1) [rectangle,draw] {Chapitre~\ref{chapter:4}\\Deux modèles MAB\\pour les réseaux IoT};
        \node[align=center] (chapter5) at (0,1) [rectangle,draw] {Chapitre~\ref{chapter:5}\\Modèle MAB\\Multi-joueurs};
        \node[align=center] (chapter6) at (2.5,1) [rectangle,draw] {Chapitre~\ref{chapter:6}\\Modèle MAB\\Non-stationnaire};
        \node[align=center] (conclusion) at (0,-0.25) [rectangle,draw] {Chapitre~\ref{chapter:conclusion}\\Conclusion générale};
        %
        \draw [color=black,thick,->] (introduction) to (chapter2);
        \draw [color=black,thick,<->] (chapter2) to (chapter3);
        \draw [color=black,thick,->] (chapter2) to (chapter4);
        \draw [color=black,thick,->] (chapter2) to (chapter5);
        \draw [color=black,densely dotted,<->]   (chapter4) to (chapter5);
        \draw [color=black,densely dotted,->] -| (chapter3) to (chapter6);
        \draw [color=black,densely dotted,<->]   (chapter5) to (chapter6);
        \draw [color=black,thick,->] (chapter2) to (chapter6);
        \draw [color=black,thick,->] (chapter4) to (conclusion);
        \draw [color=black,thick,->] (chapter5) to (conclusion);
        \draw [color=black,thick,->] (chapter6) to (conclusion);
        %
    \end{tikzpicture}
    }
    \caption[Organisation de la thèse : une carte de lecture]{Organisation de la thèse : une carte de lecture. Tout chemin contenant le Chapitre~\ref{chapter:1}, le Chapitre~\ref{chapter:2}, au moins un des trois Chapitres~\ref{chapter:4}, \ref{chapter:5} et \ref{chapter:6}, et la Conclusion est une façon valide de lire cette thèse.}
    \label{fig:1:organization_fr}
\end{figure}

\TODOL{Finir de traduire}
L'ordre de lecture du manuscrit peut être n'importe quel chemin descendant entre l'introduction dans le chapitre actuel~\ref{chapter:1}, et la conclusion dans le dernier chapitre~\ref{chapter:conclusion}.
Comme le montre le graphique de la Figure~\ref{fig:1:organization_fr} ci-dessus,
la thèse est organisée en deux parties :

### 2019-06-05 2-Chapters/2-Chapter/chapter2.tex
Fixed typos in some references

% Just bounded, in $[a,b]$ but without loss of generality in $[0,1]$ etc.
\paragraph{The normalization trick.}\label{par:2:normalizationTrick.}
%
If the player knows that reward are bounded in an interval $[a,b]$, and if she knows $a$ and $b$ (for $a<b$), then with no loss of generality we can restrict to the interval $[0,1]$, as if $r\in[a,b]$, the player can instead consider the normalized reward $r' = $ that lies in $[0,1]$.
Note that this ``normalization trick'' is implemented for any policy in our library SMPyBandits, with the \texttt{lower} and \texttt{amplitude} optional arguments, respectively representing $a$ the lower-bound on rewards and $b-a$ the length of the interval of possible values of rewards.

@@

As illustrated by the two previous examples, an efficient strategy needs to solve the trade-off between exploration and exploitation.
The two following solutions both consist in splitting the $T$ time steps into $T_0$ steps of exploration and $T-T_0$ steps of exploitation.
%  either in an alternative way or in a fixed way (``explore then exploit'').
They are detailed in Algorithm~\ref{algo:2:naiveStrategies}.

@@

% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
    \begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(){$t = 1, 2, \dots, T$}{
            \uIf{\textcolor{deeppurple}{$\varepsilon$-greedy}}{
                Sample a value uniformly in $[0,1]$: $U_t\sim\cU([0,1])$\\
                \uIf{\textcolor{deeppurple}{$U_t < \varepsilon$ (\ie, with probability $\varepsilon$)}}{
                    Play uniformly at random $A(t) \in \cU(\{1,\dots,K\})$
                    \tcp*[f]{Explore}
                }
                \uElse{
                    Play uniformly among the arms of maximal empirical mean: $A(t) \in \cU(\arg\max\limits_{1\leq k \leq K} \widehat{\mu_k}(t))$
                    \tcp*[f]{or Exploit}
                }
            }
            \uIf{\textcolor{deepgold}{Explore-then-Exploit}}{
                \uIf{$t \leq \textcolor{deepgold}{T_0}$}{
                    Play sequentially $A(t) \in 1 + (t \mod K)$
                    \tcp*[f]{Explore}
                }
                \uElse{
                    \uIf{$t = \textcolor{deepgold}{T_0}$}{
                        Pick one arm $A(\textcolor{deepgold}{T_0})=k$ uniformly among the arms of maximal empirical mean: \textcolor{deepgold}{$A(t) \in \cU(\arg\max\limits_{1\leq k \leq K} \widehat{\mu_k}(\textcolor{deepgold}{T_0}))$}
                        \tcp*[f]{then Exploit}
                    }
                    Play the same arm $A(t) = A(\textcolor{deepgold}{T_0})$
                }
            }
            Observe a reward $r_{A(t)}(t)$, and update $X_k(t)$, $N_k(t)$ and $\widehat{\mu_k}(t)$
		}
		\caption{Simple efficient strategies: \textcolor{deeppurple}{$\varepsilon$-greedy} and \textcolor{deepgold}{Explore-then-Exploit}.}
		\label{algo:2:naiveStrategies}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

@@

\begin{theorem}[Regret bound for $\UCB_1$]\label{thm:2:UCBregretBound}
    For simplicity, we restrict to $\alpha=2$.
    For any instance $I$ with $K$ arms with bounded distributions in $[0,1]$,
    of means $\mu_1,\dots,\mu_k$, and any horizon $T>K$,
    the $\UCB_1$ algorithm verifies the following finite-time regret bound
    \begin{equation}\label{eq:2:UCBregretBound}
        R_T^{\cA}(I) \leq \left(\sum_{k: \Delta_k>0} \frac{8}{\Delta_k}\right) \log(T) + \sum_{k: \Delta_k>0} \Delta_k \left(1 + \frac{\pi^2}{3}\right)
        = \bigO{ \left(\sum_{k: \Delta_k>0} \frac{8}{\Delta_k}\right) \log(T)}.
    \end{equation}
\end{theorem}
%
\begin{proof}\label{proof:2:UCBregretBound}
    \TODOL{J'aimerai bien rajouter des couleurs sur $k$ et $k^*$ partout plus bas ?}
    As we focus on bounded distributions in $[0,1]$, for which Hoeffding's inequality gives \eqref{eq:2:HoeffdingInequality}, as stated above in Lemma~\ref{lem:2:HoeffdingInequality}.
    Thanks to the regret decomposition from Lemma~\ref{lem:2:RegretDecomposition}, we can focus on bounding $\E[N_k(T)]$ for any sub-optimal arm $k$.\\
    %
    \indent
    The first trick consists in writing the following, which is true for any $n \geq 1$:
    \begin{equation}\label{eq:2:trickInUCBProof}
        N_k(T) = 1 + \sum_{t=K+1}^T \indic(A(t) = k)
        \leq n + \sum_{t=n}^T \indic(A(t) = k, N_k(t) \geq n).
    \end{equation}

>>>>

% Just bounded, in $[a,b]$ but without loss of generality in $[0,1]$ etc.
\paragraph{The normalization trick.}
\label{par:2:normalizationTrick}
%
If the player knows that reward are bounded in an interval $[a,b]$, and if she knows $a$ and $b$ (for $a<b$), then with no loss of generality we can restrict to the interval $[0,1]$, as if $r\in[a,b]$, the player can instead consider the normalized reward $r' = $ that lies in $[0,1]$.
Note that this ``normalization trick'' is implemented for any policy in our library SMPyBandits, with the \texttt{lower} and \texttt{amplitude} optional arguments, respectively representing $a$ the lower-bound on rewards and $b-a$ the length of the interval of possible values of rewards.

@@

As illustrated by the two previous examples, an efficient strategy needs to solve the trade-off between exploration and exploitation.
The two following solutions both consist in splitting the $T$ time steps into $T_0$ steps of exploration and $T-T_0$ steps of exploitation.
%  either in an alternative way or in a fixed way (``explore then exploit'').
They are detailed in Algorithm~\ref{algo:2:simpleStrategies}.

@@

% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
    \begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(){$t = 1, 2, \dots, T$}{
            \uIf{\textcolor{deeppurple}{$\varepsilon$-greedy}}{
                Sample a value uniformly in $[0,1]$: $U_t\sim\cU([0,1])$\\
                \uIf{\textcolor{deeppurple}{$U_t < \varepsilon$ (\ie, with probability $\varepsilon$)}}{
                    Play uniformly at random $A(t) \in \cU(\{1,\dots,K\})$
                    \tcp*[f]{Explore}
                }
                \uElse{
                    Play uniformly among the arms of maximal empirical mean: $A(t) \in \cU(\arg\max\limits_{1\leq k \leq K} \widehat{\mu_k}(t))$
                    \tcp*[f]{or Exploit}
                }
            }
            \uIf{\textcolor{deepgold}{Explore-then-Exploit}}{
                \uIf{$t \leq \textcolor{deepgold}{T_0}$}{
                    Play sequentially $A(t) \in 1 + (t \mod K)$
                    \tcp*[f]{Explore}
                }
                \uElse{
                    \uIf{$t = \textcolor{deepgold}{T_0}$}{
                        Pick one arm $A(\textcolor{deepgold}{T_0})=k$ uniformly among the arms of maximal empirical mean: \textcolor{deepgold}{$A(t) \in \cU(\arg\max\limits_{1\leq k \leq K} \widehat{\mu_k}(\textcolor{deepgold}{T_0}))$}
                        \tcp*[f]{then Exploit}
                    }
                    Play the same arm $A(t) = A(\textcolor{deepgold}{T_0})$
                }
            }
            Observe a reward $r_{A(t)}(t)$, and update $X_k(t)$, $N_k(t)$ and $\widehat{\mu_k}(t)$
		}
		\caption{Simple efficient strategies: \textcolor{deeppurple}{$\varepsilon$-greedy} and \textcolor{deepgold}{Explore-then-Exploit}.}
		\label{algo:2:simpleStrategies}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

@@

\begin{theorem}[Regret bound for $\UCB_1$]\label{thm:2:UCBregretBound}
    For simplicity, we restrict to $\alpha=2$.
    For any instance $I$ with $K$ arms with bounded distributions in $[0,1]$,
    of means $\mu_1,\dots,\mu_k$, and any horizon $T>K$,
    the $\UCB_1$ algorithm verifies the following finite-time regret bound
    \begin{equation}\label{eq:2:UCBregretBound}
        R_T^{\cA}(I) \leq \left(\sum_{k: \Delta_k>0} \frac{8}{\Delta_k}\right) \log(T) + \sum_{k: \Delta_k>0} \Delta_k \left(1 + \frac{\pi^2}{3}\right)
        = \bigO{ \left(\sum_{k: \Delta_k>0} \frac{8}{\Delta_k}\right) \log(T)}.
    \end{equation}
\end{theorem}
%
\begin{proof}\label{proof:2:UCBregretBound}
    \TODOL{J'aimerai bien rajouter des couleurs sur $k$ et $k^*$ partout plus bas ? si c'est bien fait, ça peut aider la lecture, par exemple \textcolor{darkgreen}{$k$} et \textcolor{gold}{$k^*$}.}
    %
    As we focus on bounded distributions in $[0,1]$, for which Hoeffding's inequality gives \eqref{eq:2:HoeffdingInequality}, as stated above in Lemma~\ref{lem:2:HoeffdingInequality}.
    Thanks to the regret decomposition from Lemma~\ref{lem:2:RegretDecomposition}, we can focus on bounding $\E[N_k(T)]$ for any sub-optimal arm $k$.\\
    %
    \indent
    The first trick consists in writing the following, which is true for any $n \geq 1$:
    \begin{equation}\label{eq:2:trickInUCBProof}
        N_k(T) = 1 + \sum_{t=K+1}^T \indic(A(t) = k)
        \leq n + \sum_{t=n}^T \indic(A(t) = k, N_k(t) \geq n).
    \end{equation}

### 2019-06-05 2-Chapters/4-Chapter/CrownCom_17.git/IoT_slotted.tex
Changed \paragraph{XXX} to \paragraph{XXX.}
Fixed typos in some references

\paragraph{Outline}
%
The rest of this section is organized as follows. The system model is introduced in Section~\ref{sub:41:systemModel}. The reference policies are described in \ref{sub:41:threeReferencePolicies}, and our proposed sequential learning policies are introduced in \ref{sub:41:sequentialPolicies}.
Numerical results are presented in \ref{sub:41:numericalResults}.

@@

\paragraph{One gateway and many devices}
%
We consider the system model presented in Figure~\ref{fig:41:system_model1}, where a set of devices all sends uplink packets to a (unique) network gateway.
Messages can be sent in a fixed number $K\geq2$ of wireless channels,
that are assumed to be orthogonal, that is, a message in channel $k$ will not cause collision to a message in another channel $j \neq i$.
% (\eg, $K=4$ or $K=16$ for instance), which are centered indifferently around the $433.5\;\mathrm{MHz}$ or the $877\;\mathrm{MHz}$ ISM bands (as it is chosen for the implementation presented in Section~\ref{sec:4:gnuradio}).
The communication between IoT devices and this gateway is done through a simple pure acknowledged ALOHA-based protocol where devices transmit uplink packets of fixed duration whenever they want.
We mean here by ALOHA-based that devices do not use sensing, and we do not consider retransmissions of packets (see Section~\ref{sec:4:retransmissions} below where we remove this last hypothesis).
%
% Note that this choice of number of channels and of frequency bands is for instance the one of the current LoRa standard in Europe, but of course the precise choice does not really matter, and future standards can choose different parameters.

@@

\paragraph{Slotted protocol}
%
For simplicity, and for consistency with the rest of this thesis, we only consider a discrete protocol (\ie, slotted).
As illustrated in Figure \ref{fig:41:protocol}, we suppose a slotted protocol, in both time and frequency.
All devices share a synchronized time, and know in advance $K$, the finite number of available RF channels.
In each time slot, the devices try to send packets to the unique gateway, which listens continuously to all channels, following an ALOHA-based communication, with no sensing.
Each time slot is divided in two parts: first for uplink communications in which data packets are sent by end-devices to the gateway.
In one channel, if only one packet is sent in this part of the slot, the gateway can decode it and sends an acknowledgement to the device in the second part (on the same channel).
If two or more devices send an uplink packet in the same slot, the uplink packets collide, and the acknowledgement \Ack{} is not transmitted. In this case, we say that there is a \emph{collision} in this time slot.
In other words, if the gateway cannot successfully decode the incoming (uplink) messages because they were corrupted by collisions, it does not send any \Ack{} back.
This way, no collision can occur on the downlink messages, easing the analysis of collisions.

@@

\paragraph{Static vs dynamic devices}
%
We make the following assumptions on the network.
We assume that there are two types of end-devices in the network:
\begin{itemize}
    \item
    \emph{Static} end-devices are identical, and each of them uses only \emph{one} channel to communicate with the gateway, with no loss of generality.
    %  if devices random access the channels.
    Their choice is assumed to be fixed in time (\ie, stationary). The traffic generated by these devices is considered as an interfering traffic for other devices.
    \item
    \emph{Dynamic} (or \emph{smart}) end-devices can use all the available channels, by quickly changing their communication channel at any time slot, following a Machine Learning policy.
    For that end, they use their history of past communication successes or failures they experienced in each channel, to learn about channel availability.
    We also assume that the dynamic end-devices can run simple embedded decision making algorithms, and have limited but reasonable computing as well as storage capacities.
    Note that of course we assume a limited storage capacity, so no device stores the full history of communication successes and failures, but they only store average rates (which can be stored using one $32$-bit IEEE 754 float number for each of the $K$ channels).
    Our goal is to propose a learning algorithm than can be implemented by each dynamic device, in a decentralized and independent manner, to improve their successful communication rate automatically.
\end{itemize}

@@

\paragraph{Random emission patterns}
%
We suppose that all devices follow the same emission pattern, being fixed in time, and we choose to model it as a simple Bernoulli process:
all devices have the same probability to send a packet in any (discrete) temporal slot, and we denote $p \in (0, 1)$ this probability\footnote{~In the experiments below, $p$ is about $10^{-3}$, because in a crowded network $p$ should be smaller than $K / (S + D)$ for all devices to communicate successfully (in average).}.
The parameter $p$ essentially controls the frequency of communication for each device, once the time scale is fixed, and $1/p$ is proportional to the \emph{duty cycle}.
% (\ie, real time during two messages)
For instance, for IoT devices sending messages with a duration of $1$ second and possibly at every second, then on a daily basis we have $p = 1 / (12 \times 60 \times 60) \simeq 1.5 \times 10^{-5}$.

\paragraph{Some assumptions on the network occupation}
%
We focus on \emph{dense networks}, in which the number of devices $S + D$ is very large compared to $K$ (for instance, about $1000$ to $100000$, while $K$ is about $4$ to $256$).
As this problem is only interesting if devices are able to communicate reasonably efficiently with the gateway, we assume that devices only communicate occasionally, \ie, with a low \emph{duty cycle}, as it is always considered for IoT.
Note that even unlicensed bands have such limitations, as for instance this is a strict requirement for any device using the $867\;\mathrm{MHz}$ ISM band in Europe (and its regulation also enforces a maximum transmission power, but we do not consider power transmission in this work).
%
We prefer this choice rather than non-crowded networks, \ie, where $S + D \leq K$, as the former makes more sense for IoT networks.
% \footnote{~Here, $K$ will typically be about $10$ different radio channels, usually in the same RF band and with separate mean carrier frequencies, and the number of devices will be of the order of $1000$ to $10000$.}.
Indeed, imagine if there were only $K=4$ channels being occupied by $D+S = 100$ devices, each communicating with a high rate of $p=1/10$, with a non-zero occupancy in each channel (\ie, $\forall k, S_k > 0$). Then, almost all time slots will lead to collisions in all channels, for a uniformly random access scheme, and thus the network efficiency (\ie, successful communication rate) will be so close to zero that using learning algorithms cannot improve much.
Such scenario is not our target of interest, and thus we prefer to only consider feasible scenarios where $p$ is smaller than $K/(S+D)$, in order to have an average of active devices not larger than the number of channels.

\paragraph{MAB model}
%
We consider the network from the point of view of one dynamic device.
Following a Bernoulli emission process, every times a dynamic device has to communicate with the gateway,
it has to choose one channel (at each transmission $t \geq 1, t \in \mathbb{N}$), denoted as $A(t) = k \in[K]$.
Then, the dynamic device waits in this channel $A(t)$ for an acknowledgement sent by the gateway, during a certain period (\eg, $5$ seconds for the example considered above).
Before sending another message (\ie, at time $t+1$), the dynamic device knows if it received or not this \Ack{} message.
%
For this reason, selecting channel (or arm) $k$ at time $t$ yields a (random) feedback, called a (binary) \emph{reward}, $r_k(t) \in \{0,1\}$, being $0$ if no \Ack{} was received before the next message, or $1$ if \Ack{} was successfully received.
The goal of the dynamic device is to minimize its packet loss ratio, or equivalently, to maximize its number of successful transmissions, or its cumulative reward,
$r_{1 \dots T} := \sum_{t = 1}^T r_{A(t)}(t),$
which is a usual objective in MAB problems.

@@

\paragraph{A bandit model for IoT}

@@

\paragraph{Two algorithms}

@@

\paragraph{\emph{Adversarial} bandit algorithms?}

@@

\begin{figure}[!b]
    \centering
    \subfloat[10\% of dynamic devices]{\includegraphics[width=0.52\textwidth]{10intelligent.eps}
    \label{fig:41:10intelligent}}
    %
    \subfloat[30\% of dynamic devices]{\includegraphics[width=0.52\textwidth]{30intelligent.eps}
    \label{fig:41:30intelligent}}
    %
    \vspace*{-10pt}
    \subfloat[50\% of dynamic devices]{\includegraphics[width=0.52\textwidth]{50intelligent.eps}
    \label{fig:41:50intelligent}}
    %
    \subfloat[100\% of dynamic devices]{\includegraphics[width=0.52\textwidth]{100intelligent.eps}
    \label{fig:41:100intelligent}}
    % FIXME use a shorter caption for the list of figures in Appendix of the thesis: it takes lots of lines to have these (a)...(d) shown in list of figures
    \caption[Performance of two MAB algorithms (\UCB{} and Thompson Sampling), compared to extreme reference policies without learning or oracle knowledge, when the proportion of dynamic end-devices in the network increases, from $10\%$ to $100\%$.]{Performance of two MAB algorithms (\UCB{} and Thompson Sampling), compared to extreme reference policies without learning or oracle knowledge, when the proportion of dynamic end-devices in the network increases, from $10\%$ to $100\%$.}
    \label{fig:41:from10to100}
    \vspace*{-10pt}
\end{figure}

\paragraph{First results}

@@

\paragraph{Successful transmissions rate as a function of the dynamic devices}

@@

\paragraph{A safety check}

@@

\paragraph{Note on the simulation code}

>>>>

\paragraph{Outline.}
%
The rest of this section is organized as follows. The system model is introduced in Section~\ref{sub:41:systemModel}. The reference policies are described in \ref{sub:41:threeReferencePolicies}, and our proposed sequential learning policies are introduced in \ref{sub:41:sequentialPolicies}.
Numerical results are presented in \ref{sub:41:numericalResults}.

@@

\paragraph{One gateway and many devices.}
%
We consider the system model presented in Figure~\ref{fig:41:system_model1}, where a set of devices all sends uplink packets to a (unique) network gateway.
Messages can be sent in a fixed number $K\geq2$ of wireless channels,
that are assumed to be orthogonal, that is, a message in channel $k$ will not cause collision to a message in another channel $j \neq i$.
% (\eg, $K=4$ or $K=16$ for instance), which are centered indifferently around the $433.5\;\mathrm{MHz}$ or the $877\;\mathrm{MHz}$ ISM bands (as it is chosen for the implementation presented in Section~\ref{sec:4:gnuradio}).
The communication between IoT devices and this gateway is done through a simple pure acknowledged ALOHA-based protocol where devices transmit uplink packets of fixed duration whenever they want.
We mean here by ALOHA-based that devices do not use sensing, and we do not consider retransmissions of packets (see Section~\ref{sec:4:retransmissions} below where we remove this last hypothesis).
%
% Note that this choice of number of channels and of frequency bands is for instance the one of the current LoRa standard in Europe, but of course the precise choice does not really matter, and future standards can choose different parameters.

@@

\paragraph{Slotted protocol.}
%
For simplicity, and for consistency with the rest of this thesis, we only consider a discrete protocol (\ie, slotted).
As illustrated in Figure \ref{fig:41:protocol}, we suppose a slotted protocol, in both time and frequency.
All devices share a synchronized time, and know in advance $K$, the finite number of available RF channels.
In each time slot, the devices try to send packets to the unique gateway, which listens continuously to all channels, following an ALOHA-based communication, with no sensing.
Each time slot is divided in two parts: first for uplink communications in which data packets are sent by end-devices to the gateway.
In one channel, if only one packet is sent in this part of the slot, the gateway can decode it and sends an acknowledgement to the device in the second part (on the same channel).
If two or more devices send an uplink packet in the same slot, the uplink packets collide, and the acknowledgement \Ack{} is not transmitted. In this case, we say that there is a \emph{collision} in this time slot.
In other words, if the gateway cannot successfully decode the incoming (uplink) messages because they were corrupted by collisions, it does not send any \Ack{} back.
This way, no collision can occur on the downlink messages, easing the analysis of collisions.

@@

\paragraph{Static vs dynamic devices.}
%
We make the following assumptions on the network.
We assume that there are two types of end-devices in the network:
\begin{itemize}
    \item
    \emph{Static} end-devices are identical, and each of them uses only \emph{one} channel to communicate with the gateway, with no loss of generality.
    %  if devices random access the channels.
    Their choice is assumed to be fixed in time (\ie, stationary). The traffic generated by these devices is considered as an interfering traffic for other devices.
    \item
    \emph{Dynamic} (or \emph{smart}) end-devices can use all the available channels, by quickly changing their communication channel at any time slot, following a Machine Learning policy.
    For that end, they use their history of past communication successes or failures they experienced in each channel, to learn about channel availability.
    We also assume that the dynamic end-devices can run simple embedded decision making algorithms, and have limited but reasonable computing as well as storage capacities.
    Note that of course we assume a limited storage capacity, so no device stores the full history of communication successes and failures, but they only store average rates (which can be stored using one $32$-bit IEEE 754 float number for each of the $K$ channels).
    Our goal is to propose a learning algorithm than can be implemented by each dynamic device, in a decentralized and independent manner, to improve their successful communication rate automatically.
\end{itemize}

@@

\paragraph{Random emission patterns.}
%
We suppose that all devices follow the same emission pattern, being fixed in time, and we choose to model it as a simple Bernoulli process:
all devices have the same probability to send a packet in any (discrete) temporal slot, and we denote $p \in (0, 1)$ this probability\footnote{~In the experiments below, $p$ is about $10^{-3}$, because in a crowded network $p$ should be smaller than $K / (S + D)$ for all devices to communicate successfully (in average).}.
The parameter $p$ essentially controls the frequency of communication for each device, once the time scale is fixed, and $1/p$ is proportional to the \emph{duty cycle}.
% (\ie, real time during two messages)
For instance, for IoT devices sending messages with a duration of $1$ second and possibly at every second, then on a daily basis we have $p = 1 / (12 \times 60 \times 60) \simeq 1.5 \times 10^{-5}$.

\paragraph{Some assumptions on the network occupation.}
%
We focus on \emph{dense networks}, in which the number of devices $S + D$ is very large compared to $K$ (for instance, about $1000$ to $100000$, while $K$ is about $4$ to $256$).
As this problem is only interesting if devices are able to communicate reasonably efficiently with the gateway, we assume that devices only communicate occasionally, \ie, with a low \emph{duty cycle}, as it is always considered for IoT.
Note that even unlicensed bands have such limitations, as for instance this is a strict requirement for any device using the $867\;\mathrm{MHz}$ ISM band in Europe (and its regulation also enforces a maximum transmission power, but we do not consider power transmission in this work).
%
We prefer this choice rather than non-crowded networks, \ie, where $S + D \leq K$, as the former makes more sense for IoT networks.
% \footnote{~Here, $K$ will typically be about $10$ different radio channels, usually in the same RF band and with separate mean carrier frequencies, and the number of devices will be of the order of $1000$ to $10000$.}.
Indeed, imagine if there were only $K=4$ channels being occupied by $D+S = 100$ devices, each communicating with a high rate of $p=1/10$, with a non-zero occupancy in each channel (\ie, $\forall k, S_k > 0$). Then, almost all time slots will lead to collisions in all channels, for a uniformly random access scheme, and thus the network efficiency (\ie, successful communication rate) will be so close to zero that using learning algorithms cannot improve much.
Such scenario is not our target of interest, and thus we prefer to only consider feasible scenarios where $p$ is smaller than $K/(S+D)$, in order to have an average of active devices not larger than the number of channels.

\paragraph{MAB model.}
%
We consider the network from the point of view of one dynamic device.
Following a Bernoulli emission process, every times a dynamic device has to communicate with the gateway,
it has to choose one channel (at each transmission $t \geq 1, t \in \mathbb{N}$), denoted as $A(t) = k \in[K]$.
Then, the dynamic device waits in this channel $A(t)$ for an acknowledgement sent by the gateway, during a certain period (\eg, $5$ seconds for the example considered above).
Before sending another message (\ie, at time $t+1$), the dynamic device knows if it received or not this \Ack{} message.
%
For this reason, selecting channel (or arm) $k$ at time $t$ yields a (random) feedback, called a (binary) \emph{reward}, $r_k(t) \in \{0,1\}$, being $0$ if no \Ack{} was received before the next message, or $1$ if \Ack{} was successfully received.
The goal of the dynamic device is to minimize its packet loss ratio, or equivalently, to maximize its number of successful transmissions, or its cumulative reward,
$r_{1 \dots T} := \sum_{t = 1}^T r_{A(t)}(t),$
which is a usual objective in MAB problems.

@@

\paragraph{A bandit model for IoT.}

@@

\paragraph{Two algorithms.}

@@

\paragraph{\emph{Adversarial} bandit algorithms?.}

@@

\begin{figure}[!b]
    \centering
    \subfloat[10\% of dynamic devices]{\includegraphics[width=0.52\textwidth]{10intelligent.eps}
    \label{fig:41:10intelligent}}
    %
    \subfloat[30\% of dynamic devices]{\includegraphics[width=0.52\textwidth]{30intelligent.eps}
    \label{fig:41:30intelligent}}
    %
    \vspace*{-10pt}
    \subfloat[50\% of dynamic devices]{\includegraphics[width=0.52\textwidth]{50intelligent.eps}
    \label{fig:41:50intelligent}}
    %
    \subfloat[100\% of dynamic devices]{\includegraphics[width=0.52\textwidth]{100intelligent.eps}
    \label{fig:41:100intelligent}}
    % FIXME use a shorter caption for the list of figures in Appendix of the thesis: it takes lots of lines to have these (a)...(d) shown in list of figures
    \caption[XXX Performance of two MAB algorithms (\UCB{} and Thompson Sampling), compared to extreme reference policies without learning or oracle knowledge, when the proportion of dynamic end-devices in the network increases, from $10\%$ to $100\%$.]{Performance of two MAB algorithms (\UCB{} and Thompson Sampling), compared to extreme reference policies without learning or oracle knowledge, when the proportion of dynamic end-devices in the network increases, from $10\%$ to $100\%$.}
    \label{fig:41:from10to100}
    \vspace*{-10pt}
\end{figure}

\paragraph{First results.}

@@

\paragraph{Successful transmissions rate as a function of the dynamic devices.}

@@

\paragraph{A safety check.}

@@

\paragraph{Note on the simulation code.}
